Modelling the emergence of whisker barrels

### 2020-07-11 paper/briefpaper/brief.tex
minor edits mainly to fix small typos and minor grammatical issues

\cite{karbowski_model_2004} developed a reaction-diffusion style model of how
extrinsic signalling gradients can constrain the emergence of distinct fields
from intrinsic cortical dynamics. Their model defines how the \cmnt{fraction
  of occupied synapses} $c_i(x,t)$ and \cmnt{the density of} axon branches
$a_i(x,t)$ interact at time $t$, along a 1D anterior-posterior axis $x$, for
$N$ thalamocortical projections indexed by $i$. The model was derived from the
assumption that the rates at which $a_i$ and $c_i$ grow are reciprocally
coupled. Extending the original 1D model to simulate arealisation on a 2D
cortical sheet, we use $a_i(\mb{x},t)$ and $c_i(\mb{x},t)$, and model
synaptogenesis as
%
\begin{equation} \label{eq:dc}
\frac{\partial c_i}{\partial t} =-\alpha c_i +\beta  \left(1 - \sum_{j=1}^{N} c_{j}\right)[a_i]^k.
\end{equation}
%
Accordingly, where the total \cmnt{fraction} of synaptic connections sums to one,
connections decay at rate $\alpha$. Otherwise, \cmnt{$c_i(\mb{x},t)$} increases
non-linearly ($k>1$) with the density of axon branching. Axon branching is
modelled as
%
\begin{equation} \label{eq:da}
\frac{\partial a_i}{\partial t} = \nabla\vcdot\left(D \nabla a_i-a_i\sum_{j=1}^{M} \gamma_{i,j}\nabla \rho_j(\mb{x}) \MPtwo{+ \chi_i}\right) - \frac{\partial c_i}{\partial t}.
\end{equation}
%
The first term on the right describes the divergence (indicated by
$\nabla\vcdot$) of the quantity in parentheses, which is referred to as the
`flux' of axonal branching. The flux represents diffusion across the cortical
sheet, at rate $D$, and the influence of $M$ molecular signalling fields,
$\rho(\mb{x})$. The influence of a given field (indexed by $j$) on a given
thalamic projection (indexed by $i$), is determined by $\gamma_{i,j}$, which
may be positive or negative in order that axons may branch in the direction of
either higher or lower concentrations. Note that computing the divergence in
simulation requires cells on the cortical sheet to communicate with immedately
adjacent cells only (see \emph{Materials \& Methods}). Here $\chi_i=0$ is a
placeholder. The second term on the right represents the coupling between axon
branching and synaptogenesis \cmnt{and implies the assumption that the spatial
  distribution of synaptic density across the cortical sheet is broadly
  homogeneous. This makes it possible to relate the quantity $c_i$ to
  connection density}.

@@

% New section to describe and introduce the metrics
\metrics{First, we used an algorithm introduced by Honda to measure the
  discrepancy of each barrel shape from a Dirichletform shape}
\citep{honda_geometrical_1983}.  \metrics{Low overall values of this
  \emph{Honda-}$\delta$ metric obtained from simulated barrels indicate a
  close correspondence of the simulated barrel field with a Voronoi
  tesselation, and thus with a biological barrel field.
  (In mouse, $\delta\approx0.054$} \citep{senft_mouse_1991}\metrics{;
  our analysis of data from} \cite{zheng_signal_2001} \metrics{ indicates that
  the value for rat is similar). For the tesselation
  that is overlaid on the real barrel field in Fig.\,\ref{fig:main}A,
  $\delta=0.025$, and a reduction in $\delta$ in the example simulation over
  time confirmed that an equivalent `good' Voronoi pattern can emerge
  within $\approx$~20000
  iterations (Fig.\,\ref{fig:main}D, red circles). Second, we devised a
  \emph{pattern difference} measure that is sensitive to deviations in the
  component shapes and overall topographic registration between two
  tesselations, $\eta$, and we used this measure to compare the simulated
  barrel fields to the real barrel field from which the boundary shape applied
  to the simulation was obtained (see \emph{Materials \& Methods} for
  details). A similar reduction in $\eta$ in the development of the example
  simulation confirmed that the shapes and arrangement of emergent connection
  fields came to match those of the real barrel field by around 20000
  iterations (Fig.\,\ref{fig:main}D, black squares). Third, we measured the
  \emph{connection selectivity} at each location on the cortical sheet, as the
  connection density of the most dense projection divided by the sum over all
  projection densities, $\omega$. The connection selectivity increased as the
  barrel map self-organized in the example simulation (Fig.\,\ref{fig:main}D,
  grey hexagons), and the selectivity became concentrated in regions
  overlapping with the emergent barrel centers (Fig.\,\ref{fig:main}E).}

\metrics{Against these three metrics we are also able to characterise the
  robustness of self-organization to the model parameters, and to investigate
  the sensitivity of the model to variation in its inputs.}

@@

\MPthreePar{Fig.\,\ref{fig:paramsweep}A shows values of $\delta$, $\eta$ and
  $\omega$ obtained after 30000 iterations, from 216 independent simulations,
  each representing a unique combination of the model parameters $D$,
  $\epsilon$, and the ratio $\alpha/\beta$. First we observe that
  self-organization is highly robust to the ratio $\alpha/\beta$, across five
  orders of magnitude, with respect to all three metrics. Second, the most
  strongly Dirichletform patterns (low \emph{Honda}-$\delta$) were generated
  by simulations in which the diffusion constant $D$ and the strength of
  competition $\epsilon$ were high. Third, strongest connection selectivities,
  $\omega$, were obtained for lower values of $D$. Fourth, variation in the
  pattern difference metric, $\eta$, indicated that the alignment between real
  and simulated patterns was greatest for intermediate rates of diffusion,
  $D\approx0.5$. Together these results indicate that when competition is
  strong, the rate of diffusion determines a trade-off such that fields emerge
  to be barrel-shaped when diffusion is fast and they emerge to be more
  selectively innervated when diffusion is slow.}

@@

% This figure is text width
\begin{center}
  \includegraphics[width=\linewidth]{./Fig4.png} \captionof{figure}{
    \MPthreePred{Simulating altered barrel development.  Guidance fields
      (\textbf{A}) and emergent barrel pattern (\textbf{B}) in a Fgf8
      misexpression experiment
      (c.f.~\citealp{assimacopoulos_fibroblast_2012}), simulated by reflecting
      $\rho_1$ from Fig.\,\ref{fig:main}A at the join of the original boundary
      with its mirror. All other model parameters match those in
      Fig.\,\ref{fig:main}C. \textbf{C} Simulating whisker trimming by
      reducing the competitiveness, $\epsilon$, of the one projection. For C3
      only, $\epsilon$ was mutiplied by $m\in(0,1)$. The pattern shown is that
      formed after 30000 steps with $m=0.86$, which reduces the size of the C3
      field to 65\% of its original size, matching the average barrel area
      reduction observed by }
    \citealp{kossut_effects_1992}. \MPthreePred{\textbf{D} The area of the C3
      barrel (black squares) reduces as $m$ is reduced, whereas the mean area
      of neighbouring barrels (B3, C2, C4, D2 \& D3, grey circles)
      increases. The dotted grey line indicates $m=0.86$ for comparison with
      panel C. If the $\epsilon$ value is instead reduced for \emph{all} row C
      projections (including the interstitial $\gamma$ projection), the mean
      row C barrel area is only slightly reduced (light grey triangles). In
      this case, the mean area of a simulated row C barrel at $m=0.86$ is 91\%
      of that for $m=1$.}}
  \label{fig:fgf8}
\end{center}

>>>>

\cite{karbowski_model_2004} developed a reaction-diffusion style model of how
extrinsic signalling gradients can constrain the emergence of distinct fields
from intrinsic cortical dynamics. Their model defines how the \cmnt{fraction
  of occupied synapses} $c_i(x,t)$ and \cmnt{the density of} axon branches
$a_i(x,t)$ interact at time $t$, along a 1D anterior-posterior axis $x$, for
$N$ thalamocortical projections indexed by $i$. The model was derived from the
assumption that the rates at which $a_i$ and $c_i$ grow are reciprocally
coupled. Extending the original 1D model to simulate arealisation on a 2D
cortical sheet, we use $a_i(\mb{x},t)$ and $c_i(\mb{x},t)$, and model
synaptogenesis as
%
\begin{equation} \label{eq:dc}
\frac{\partial c_i}{\partial t} =-\alpha c_i +\beta  \left(1 - \sum_{j=1}^{N} c_{j}\right)[a_i]^k.
\end{equation}
%
Accordingly, where the total \cmnt{fraction} of synaptic connections sums to one,
connections decay at rate $\alpha$. Otherwise, \cmnt{$c_i(\mb{x},t)$} increases
non-linearly ($k>1$) with the density of axon branching. Axon branching is
modelled as
%
\begin{equation} \label{eq:da}
\frac{\partial a_i}{\partial t} = \nabla\vcdot\left(D \nabla a_i-a_i\sum_{j=1}^{M} \gamma_{i,j}\nabla \rho_j(\mb{x}) \MPtwo{+ \chi_i}\right) - \frac{\partial c_i}{\partial t}.
\end{equation}
%
The first term on the right describes the divergence (indicated by
$\nabla\vcdot$) of the quantity in parentheses, which is referred to as the
`flux' of axonal branching. The flux represents diffusion across the cortical
sheet, at rate $D$, and the influence of $M$ molecular signalling fields,
$\rho(\mb{x})$. The influence of a given field (indexed by $j$) on a given
thalamic projection (indexed by $i$), is determined by $\gamma_{i,j}$, which
may be positive or negative in order that axons may branch in the direction of
either higher or lower concentrations. Note that computing the divergence in
simulation requires cells on the cortical sheet to communicate with immedately
adjacent cells only (see \emph{Materials \& Methods}). Here $\chi_i=0$ is a
placeholder. The second term on the right represents the coupling between axon
branching and synaptogenesis, \cmnt{and an assumption that the spatial
  distribution of synaptic density across the cortical sheet is broadly
  homogeneous. As such, the quantity $c_i$ can be thought of as the
  connection density}.

@@

% New section to describe and introduce the metrics
\metrics{First, we used an algorithm introduced by Honda to measure the
  discrepancy of each barrel shape from a Dirichletform shape}
\citep{honda_geometrical_1983}.  \metrics{Low overall values of this
  \emph{Honda-}$\delta$ metric obtained from simulated barrels indicate a
  close correspondence of the simulated barrel field with a Voronoi
  tesselation, and thus with a biological barrel field 
  (for mice, $\delta\approx0.054$} \citealp{senft_mouse_1991}\metrics{, and 
  our analysis of data from} \citealp{zheng_signal_2001} \metrics{ indicates that
  the value for rats is similar). For the tesselation
  that is overlaid on the real barrel field in Fig.\,\ref{fig:main}A,
  $\delta=0.025$, and a reduction in $\delta$ in the example simulation over
  time confirmed that an equivalent `good' Voronoi pattern can emerge
  within $\approx$~20000
  iterations (Fig.\,\ref{fig:main}D, red circles). Second, we devised a
  \emph{pattern difference} measure that is sensitive to deviations in the
  component shapes and overall topographic registration between two
  tesselations, $\eta$, and we used this measure to compare the simulated
  barrel fields to the real barrel field from which the boundary shape applied
  to the simulation was obtained (see \emph{Materials \& Methods} for
  details). A similar reduction in $\eta$ in the development of the example
  simulation confirmed that the shapes and arrangement of emergent connection
  fields came to match those of the real barrel field by around 20000
  iterations (Fig.\,\ref{fig:main}D, black squares). Third, we measured the
  \emph{connection selectivity} at each location on the cortical sheet, as the
  connection density of the most dense projection divided by the sum over all
  projection densities, $\omega$. The overall connection selectivity increased as the
  barrel map self-organized in the example simulation (Fig.\,\ref{fig:main}D,
  grey hexagons), and the selectivity became concentrated in regions
  overlapping with the emergent barrel centers (Fig.\,\ref{fig:main}E).}

@@

\metrics{Against these three metrics we are also able to characterise the
  robustness of self-organization to the model parameters, and to investigate
  the sensitivity of the model to variation in its inputs.} \MPthreePar{Fig.\,\ref{fig:paramsweep}A shows values of $\delta$, $\eta$ and
  $\omega$ obtained after 30000 iterations, from 216 independent simulations,
  each representing a unique combination of the model parameters $D$,
  $\epsilon$, and the ratio $\alpha/\beta$. First we observe that
  self-organization is highly robust to the ratio $\alpha/\beta$, across five
  orders of magnitude, with respect to all three metrics. Second, the most
  strongly Dirichletform patterns (low \emph{Honda}-$\delta$) were generated
  by simulations in which the diffusion constant $D$ and the strength of
  competition $\epsilon$ were high. Third, strongest overall connection selectivities,
  $\omega$, were obtained for lower values of $D$. Fourth, variation in the
  pattern difference metric, $\eta$, indicated that the alignment between real
  and simulated patterns was greatest for intermediate rates of diffusion,
  $D\approx0.5$. Together these results indicate that when competition is
  strong, the rate of diffusion determines a trade-off such that fields emerge
  to be barrel-shaped when diffusion is fast and they emerge to be more
  selectively innervated when diffusion is slow.}

@@

% This figure is text width
\begin{center}
  \includegraphics[width=\linewidth]{./Fig4.png} \captionof{figure}{
    \MPthreePred{Simulating altered barrel development.  Guidance fields
      (\textbf{A}) and emergent barrel pattern (\textbf{B}) in a Fgf8
      misexpression experiment
      (c.f.~\citealp{assimacopoulos_fibroblast_2012}), simulated by reflecting
      $\rho_1$ from Fig.\,\ref{fig:main}A at the join of the original boundary
      with its mirror. All other model parameters match those in
      Fig.\,\ref{fig:main}C. \textbf{C} Simulating whisker trimming by
      reducing the competitiveness, $\epsilon$, of one projection. For C3
      only, $\epsilon$ was mutiplied by $m\in(0,1)$. The pattern shown is that
      formed after 30000 steps with $m=0.86$, which reduces the size of the C3
      field to 65\% of its original size, matching the average barrel area
      reduction observed by }
    \cite{kossut_effects_1992}. \MPthreePred{\textbf{D} The area of the C3
      barrel (black squares) reduces as $m$ is reduced, whereas the mean area
      of neighbouring barrels (B3, C2, C4, D2 \& D3, grey circles)
      increases. The dotted grey line indicates $m=0.86$ for comparison with
      panel \textbf{C}. If the $\epsilon$ value is instead reduced for \emph{all} row C
      projections (including the interstitial $\gamma$ projection), the mean
      row C barrel area is only slightly reduced (light grey triangles). In
      this case, the mean area of a simulated row C barrel at $m=0.86$ is 91\%
      of that for $m=1$.}}
  \label{fig:fgf8}
\end{center}
